---
title: "STAT 146 US Total Lithium Carbonate Imports Analysis"
author: "Victoria Rios"
date: "12/3/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## background  
lithium carbonate is a form of clean lithium used in lithium-ion batteries, in everything from your phone to a Tesla battery. 
The recent US backed coup in Bolivia was due partially to their lithium reserves.  
The following is monthly imports of lithium carbonate in customs value ($US)^[The value of goods imported as appraised by U.S. Customs and Border Protection. This value is generally defined as the price actually paid or payable for merchandise when sold for exportation to the U.S. It excludes U.S. import duties, freight, insurance, and other charges incurred in bringing the merchandise to the U.S. (General Imports)] to US ports from January 2003 to September 2019 from usatrade.census.gov

how much moneys of lithium is coming into the united states each month?  
can we see any trends? (spikes at electric car production?)  
can we predict future without taking politics into account?  

## Cleaning section

```{r}
# for cleaning
library(tidyverse)
library(lubridate)
# the textbook
library(TSA)

# do not alter lithium.raw!
lithium.raw <- read.csv("usa_lithium_monthly_port_level_imports.csv",
                    skip = 4, header = T)

lithium <- lithium.raw %>%
  rename(CustomsValue = Customs.Value..Gen....US.) %>% # rename to something friendlier
  select(Time, CustomsValue) %>% # not the null or Total/lithum carbonate columns
  distinct() %>% # clear out the double lines
  separate(Time, into = c("Month", "Year"), convert = T) %>%
  mutate(monthnum = rep(1:12, length.out = 201)) %>% # this is janky code
  mutate(time = make_date(Year, monthnum)) %>%
  select(time, CustomsValue, Year, monthnum) # kept year & monthnum just in case

lithium.ts <- ts(lithium$CustomsValue,
                 start = c(2003, 1), end = c(2019, 9), frequency = 12)
plot(lithium.ts)
```

